---
output: github_document
bibliography: "Z:/studia/MINI/praca_magisterska/metoda_reprezentacyjna/thesis/references.bib"
references:
- id: wojciak2019
  title: Optimal Allocation in Stratified Sampling Schemes
  author:
  - family: Wojciak
    given: Wojciech
  container-title: Master's diploma thesis
  URL: 'http://home.elka.pw.edu.pl/~wwojciak/msc_optimal_allocation.pdf'
  publisher: Warsaw University of Technology
  type: thesis
  issued:
    year: 2019
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# stratallo

<!-- badges: start -->
<!-- badges: end -->

The goal of stratallo is to provide methods computing solution to classical problem in survey methodology - an optimum sample allocation problem in stratified sampling scheme with simple random sampling without replacement design in each stratum. In this context, the optimal allocation is in the classical Tschuprow-Neyman's sense [@Tschuprow1923; @Neyman1934], and it satisfies additional upper bounds restrictions imposed on sample sizes in strata. There are four different algorithms available to use, and one of them is Neyman optimal allocation applied in a recursive way [see @SarndalSwensson1993, Remark 12.7.1, p. 466]. All the algorithms are described in detail, including proofs in @wojciak2019.

## Installation

You can install the released version of stratallo from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("stratallo")
```

## Example

This is a basic example which shows you how to solve a common problem of optimal sample allocation for an example population with 4 strata.

```{r example}
library(stratallo)

N <- c(3000, 4000, 5000, 2000) # strata sizes
S <- c(48, 79, 76, 17) # standard deviations of a study variable in strata
M <- c(100, 90, 70, 80) # upper bounds constraints imposed on the sample sizes in strata
all(M <= N) 
n <- 190 # total sample size
n <= sum(M)

nopt_ <- nopt(n = n, N = N, S = S, M = M)
nopt_
sum(nopt_)

# verbose output + variance for a given allocation
nopt(n = n, N = N, S = S, M = M, verbose = TRUE, variance = TRUE)

```

# References
