---
title: "Optimum Sample Allocation in Stratified Sampling with `stratallo` Package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Optimum Sample Allocation in Stratified Sampling with `stratallo` Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
csl: "asa_modified_url.csl"
references:
- id: neyman
  type: article-journal
  title: "On the Two Different Aspects of the Representative Method: The Method of
    Stratified Sampling and the Method of Purposive Selection"
  author:
  - family: Neyman
    given: Jerzy
  container-title: Journal of the Royal Statistical Society
  volume: 97
  number: 4
  page: 558-606
  issued:
    year: 1934
- id: tschuprov
  type: article-journal
  title: On the Mathematical Expectation of the Moments of Frequency Distributions
    in the Case of Correlated Observations
  author:
  - family: Tschuprov
    given: Alexander Alexandrovich
  container-title: Metron
  volume: 2
  number: 4
  page: 461-493, 646-683
  issued:
    year: 1923
- id: wesolowski2021
  type: article-journal
  title: Optimality of the Recursive Neyman Allocation
  author:
  - family: Wesołowski
    given: Jacek
  - family: Wieczorkowski
    given: Robert
  - family: Wójciak
    given: Wojciech
  container-title: "Journal of Survey Statistics and Methodology"
  DOI: 10.1093/jssam/smab018
  URL: https://arxiv.org/abs/2105.14486
  issued:
    year: 2021
- id: wojciak2023
  type: article
  title: Another Solution of Some Optimum Allocation Problem
  author:
  - family: Wójciak
    given: Wojciech
  issued:
    year: 2023
- id: wojciak2019
  type: thesis
  title: Optimum Allocation in Stratified Sampling Schemes
  author:
  - family: Wójciak
    given: Wojciech
  container-title: MSc Thesis
  URL: 'http://home.elka.pw.edu.pl/~wwojciak/msc_optimal_allocation.pdf'
  publisher: Warsaw University of Technology
  issued:
    year: 2019
- id: sarndal
  type: book
  title: Model Assisted Survey Sampling
  author:
  - family: Sarndal
    given: Carl-Erik
  - family: Swensson
    given: Bengt
  - family: Wretman
    given: Jan
  publisher: Springer
  issued:
    year: 1993
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

\def\R{{\mathbb R}}
\def\x{\mathbf x}
\def\n{\mathbf n}
\def\D{\widetilde D}
\newcommand{\texteq}{\mathrm}

The goal of *stratallo* package is to provide implementations of the efficient 
algorithms that solve a classical problem in survey methodology - an optimum
sample allocation in stratified sampling. In this context, the classical problem
of optimum sample allocation is the Tschuprov-Neyman's sense
[@tschuprov; @neyman]. It is formulated as determination of a vector of strata
sample sizes that minimizes the variance of the *stratified $\pi$-estimator* of
the population total of a given study variable, under constraint on total sample
size. This problem can be further complemented by imposing lower or upper bounds
on sample sizes is strata.

A minor modification of the classical optimum sample allocation problem leads
to the minimum cost allocation. This problem lies in the determination of a
vector of strata sample sizes that minimizes total cost of the survey, under
assumed fixed level of the stratified $\pi$-estimator's variance. As in the case
of the classical optimum allocation, the problem of minimum cost allocation can
be complemented by imposing upper bounds on sample sizes in strata.

*stratallo* provides two **user functions**:

* `opt()`
* `optcost()`

that solve sample allocation problems briefly characterized above. In this
context, it is assumed that the variance of the stratified estimator is of the
following generic form:
$$
  V_{st}(\n) = \sum_{h=1}^{H} \frac{A_h^2}{n_h} - A_0,
$$
where $H$ denotes total number of strata, $\n = (n_h)_{h \in \{1,\ldots,H\}}$ is
the allocation vector with strata sample sizes, and population parameters $A_0$,
and $A_h > 0,\, h = 1,\ldots,H$, do not depend on the $x_h,\, h = 1,\ldots,H$.

Among stratified estimators and stratified sampling designs that jointly give
rise to a variance of the above form, is the so called stratified $\pi$
estimator of the population total with *stratified simple random sampling
without replacement* design, which is one of the most basic and commonly used
stratified sampling designs. This case yields
$A_0 = \sum_{h = 1}^H N_h S_h^2$, $A_h = N_h S_h,\, h = 1,\ldots,H$, where
$S_h$ denotes stratum standard deviation of study variable and $N_h$ is the
stratum size (see e.g. @sarndal, Result 3.7.2, p.103).

Apart from `opt()` and `optcost()`, *stratallo* provides the following
**helpers functions**:

* `var_st()`,
* `var_st_tsi()`,
* `asummary()`,
* `ran_round()`,
* `round_oric()`.

Functions `var_st()` and `var_st_tsi()` compute a value of the variance $V_{st}$.
The `var_st_tsi()` is a simple wrapper of `var_st()` that is dedicated for the
case when $A_0 = \sum_{h = 1}^H N_h S_h^2$ and $A_h = N_h S_h,\, h = 1,\ldots,H$.
`asummary()` creates a `data.frame` object with summary of the allocation.
Functions `ran_round()` and `round_oric()` are the rounding functions that can
be used to round non-integers allocations (see section Rounding, below).
The package comes with three predefined, artificial populations with 10, 507
and 969 strata. These are stored under `pop10_mM`, pop507` and `pop969` objects,
respectively.

## Minimization of the variance with `opt()` function

The `opt()` function solves the following three problems of the optimum sample
allocation, formulated in the language of mathematical optimization. User of
`opt()` can choose whether the solution computed will be for 
**Problem 1**, **Problem 2** or **Problem 3**. This is achieved with the proper
use of `m` and `M` arguments of the function. Also, if required, the inequality
constraints can be removed from the optimization problem. For more details, see
the help page for `opt()` function.

### Problem 1 (one-sided upper bounds constraints)
Given numbers $n > 0,\, A_h > 0,\, M_h > 0$, such that
$M_h \leq N_h,\, h = 1,\ldots,H$, and $n \leq \sum_{h=1}^H M_h$, 
\begin{align*}
	\underset{\x \in \R_+^H}{\texteq{minimize ~\,}} & \quad f(\x) = \sum_{h=1}^H \tfrac{A_h^2}{x_h} \\
	\texteq{subject ~ to}	& \quad \sum_{h=1}^H x_h = n \\
	& \quad x_h \leq M_h, \quad{h = 1,\ldots,H,}
\end{align*}
where $\x = (x_h)_{h \in \{1,\ldots,H\}}$.

There are four different algorithms available to use for **Problem 1**,
*RNA* (default), *SGA*, *SGAPLUS*, *COMA*. All these algorithms, except
*SGAPLUS*, are described in detail in @wesolowski2021. The *SGAPLUS* is defined
in @wojciak2019 as *Sequential Allocation (version 1)* algorithm.

#### Examples

```{r load_package}
library(stratallo)
```

Define example population

```{r pop}
N <- c(3000, 4000, 5000, 2000) # Strata sizes.
S <- c(48, 79, 76, 16) # Standard deviations of a study variable in strata.
a <- N * S
n <- 190 # Total sample size.
```

Tschuprov-Neyman allocation (no inequality constraints).

```{r opt_Neyman}
opt <- opt(n = n, a = a)
opt
sum(opt) == n
# Variance of the stratified estimator that corresponds to optimum allocation.
var_st_tsi(opt, N, S)
```

One-sided upper-bounds constraints.

```{r opt_M}
M <- c(100, 90, 70, 80) # Upper bounds imposed on the sample sizes in strata.
all(M <= N)
n <= sum(M)

# Solution to Problem 1.
opt <- opt(n = n, a = a, M = M)
opt
sum(opt) == n
all(opt <= M) # Does not violate upper-bounds constraints.
# Variance of the stratified estimator that corresponds to optimum allocation.
var_st_tsi(opt, N, S)
```

### Problem 2 (one-sided lower-bounds constraints)
Given numbers $n,\, A_h > 0,\, m_h > 0$, such that
$m_h \leq N_h,\, h = 1,\ldots,H$, and $n \geq \sum_{h=1}^H m_h$, 
\begin{align*}
	\underset{\x \in \R_+^H}{\texteq{minimize ~\,}} & \quad f(\x) = \sum_{h=1}^H \tfrac{A_h^2}{x_h} \\
	\texteq{subject ~ to} & \quad \sum_{h=1}^H x_h = n \\
	& \quad x_h \geq m_h, \quad{h = 1,\ldots,H,}
\end{align*}
where $\x = (x_h)_{h \in \{1,\ldots,H\}}$.

The optimization **Problem 2** is solved by the *LRNA* that in principle is
based on the *RNA* and it is introduced in @wojciak2023.

#### Examples

```{r opt_m}
m <- c(50, 120, 1, 2) # Lower bounds imposed on the sample sizes in strata.
n >= sum(m)

# Solution to Problem 2.
opt <- opt(n = n, a = a, m = m)
opt
sum(opt) == n
all(opt >= m) # Does not violate lower-bounds constraints.
# Variance of the stratified estimator that corresponds to optimum allocation.
var_st_tsi(opt, N, S)
```

### Problem 3 (box constraints)
Given numbers $n,\, A_h > 0,\, m_h > 0,\, M_h > 0$, such that
$m_h < M_h \leq N_h,\, h = 1,\ldots,H$, and $\sum_{h=1}^H m_h \leq n \leq \sum_{h=1}^H M_h$, 
\begin{align*}
	\underset{\x \in \R_+^H}{\texteq{minimize ~\,}}  & \quad f(\x) = \sum_{h=1}^H \tfrac{A_h^2}{x_h} \\
	\texteq{subject ~ to} & \quad \sum_{h=1}^H x_h = n \\
	& \quad m_h \leq x_h \leq M_h, \quad{h = 1,\ldots,H,}
\end{align*}
where $\x = (x_h)_{h \in \{1,\ldots,H\}}$.

The optimization **Problem 3** is solved by the *RNABOX* which is a new algorithm
proposed by the authors of this package and it will be published soon.

#### Examples

```{r opt_box}
m <- c(100, 90, 500, 50) # Lower bounds imposed on sample sizes in strata.
M <- c(300, 400, 800, 90) # Upper bounds imposed on sample sizes in strata.
n <- 1284
n >= sum(m) && n <= sum(M)

# Optimum allocation under box constraints.
opt <- opt(n = n, a = a, m = m, M = M)
opt
sum(opt) == n
all(opt >= m & opt <= M) # Does not violate any lower or upper bounds constraints.
# Variance of the stratified estimator that corresponds to optimum allocation.
var_st_tsi(opt, N, S)
```

## Minimization of the total cost with `optcost()` function

The `optcost()` function solves the following minimum total cost allocation
problem, formulated in the language of mathematical optimization.

### Problem 4
Given numbers $A_h > 0,\, c_h > 0,\, M_h > 0$, such that
$M_h \leq N_h,\, h = 1,\ldots,H$, and $V \geq \sum_{h=1}^H \tfrac{A_h^2}{M_h} - A_0$, 
\begin{align*}
	\underset{\x \in \R_+^H}{\texteq{minimize ~\,}}  & \quad c(\x) = \sum_{h=1}^H c_h x_h \\
	\texteq{subject ~ to}	& \quad \sum_{h=1}^H \tfrac{A_h^2}{x_h} - A_0 = V \\
	& \quad x_h \leq M_h, \quad{h = 1,\ldots,H,}
\end{align*}
where $\x = (x_h)_{h \in \{1,\ldots,H\}}$.

The algorithm that solves **Problem 4** is based on the *LRNA* and it is
described in @wojciak2023.

#### Examples

```{r optcost}
a <- c(3000, 4000, 5000, 2000)
a0 <- 70000
unit_costs <- c(0.5, 0.6, 0.6, 0.3) # c_h, h = 1,...4.
M <- c(100, 90, 70, 80)
V <- 1e6 # Variance constraint.
V >= sum(a^2 / M) - a0

opt <- optcost(V = V, a = a, a0 = a0, M = M, unit_costs = unit_costs)
opt
sum(a^2 / opt) - a0 == V
all(opt <= M)
```

## Rounding

*stratallo* comes with 2 functions: `ran_round()` and `round_oric()` that can
be used to round non-integer allocations.

#### Examples

```{r rounding}
m <- c(100, 90, 500, 50)
M <- c(300, 400, 800, 90)
n <- 1284

# Optimum, non-integer allocation under box constraints.
opt <- opt(n = n, a = a, m = m, M = M)
opt

opt_int <- round_oric(opt)
opt_int
```

## Installation

You can install the released version of *stratallo* package from
[CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("stratallo")
```

## Note on finite precision arithmetic

Consider the following example

```{r finit_prec1}
N <- c(3000, 4000, 5000, 2000)
S <- c(48, 79, 76, 17)
a <- N * S
n <- 190
opt <- opt(n = n, a = a) # which after simplification is (n / sum(a)) * a
opt
```

and note that

```{r finit_prec2}
sum(opt) == n
```

which results from the fact that

```{r finit_prec3}
options(digits = 22)
sum(opt)

sum((n / sum(a)) * a) == n # mathematically, it should be TRUE!
```

## References
