---
title: "Optimum Sample Allocation in Stratified Sampling Schemes with Stratallo Package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Optimum Sample Allocation in Stratified Sampling Schemes with Stratallo Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
csl: "asa_modified_url.csl"
references:
- id: neyman
  type: article-journal
  title: "On the Two Different Aspects of the Representative Method: The Method of
    Stratified Sampling and the Method of Purposive Selection"
  author:
  - family: Neyman
    given: Jerzy
  container-title: Journal of the Royal Statistical Society
  volume: 97
  number: 4
  page: 558-606
  issued:
    year: 1934
- id: tschuprov
  type: article-journal
  title: On the Mathematical Expectation of the Moments of Frequency Distributions
    in the Case of Correlated Observations
  author:
  - family: Tschuprov
    given: Alexander Alexandrovich
  container-title: Metron
  volume: 2
  number: 4
  page: 461-493, 646-683
  issued:
    year: 1923
- id: wesolowski2021
  type: article-journal
  title: Optimality of the Recursive Neyman Allocation
  author:
  - family: Wesołowski
    given: Jacek
  - family: Wieczorkowski
    given: Robert
  - family: Wójciak
    given: Wojciech
  container-title: "Journal of Survey Statistics and Methodology"
  DOI: 10.1093/jssam/smab018
  URL: https://arxiv.org/abs/2105.14486
  issued:
    year: 2021
- id: wojciak2022
  type: article
  title: Minimum Sample Size Allocation in Stratified Sampling Under Constraints on
    Variance and Strata Sample Sizes
  author:
  - family: Wójciak
    given: Wojciech
  DOI: 10.48550/arXiv.2204.04035
  issued:
    year: 2022
- id: wojciak2019
  type: thesis
  title: Optimal Allocation in Stratified Sampling Schemes
  author:
  - family: Wójciak
    given: Wojciech
  container-title: MSc Thesis
  URL: 'http://home.elka.pw.edu.pl/~wwojciak/msc_optimal_allocation.pdf'
  publisher: Warsaw University of Technology
  issued:
    year: 2019
- id: sarndal
  type: book
  title: Model Assisted Survey Sampling
  author:
  - family: Sarndal
    given: Carl-Erik
  - family: Swensson
    given: Bengt
  - family: Wretman
    given: Jan
  publisher: Springer
  issued:
    year: 1993
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

\def\W{\mathcal W}
\def\x{\mathbf x}
\def\n{\mathbf n}
\def\D{\widetilde D}
\newcommand{\texteq}{\mathrm}

The goal of *stratallo* package is to provide implementations of the efficient 
algorithms that solve a classical problem in survey methodology - an optimum
sample allocation problem in stratified sampling schemes. In this context, the
classical problem of optimum sample allocation is the Tschuprov-Neyman's sense
[@tschuprov; @neyman]. It is formulated as determination of a vector of strata
sample sizes that minimizes the variance of the $\pi$-estimator of the 
population total of a given study variable, under constraint on total sample
size. This problem can be further complemented by adding lower or upper bounds
constraints on sample sizes is strata.

A minor modification of the classical optimium sample allocation problem leads
to the minimum sample size allocation. This problems lies in the determination
of a vector of strata sample sizes that minimizes total sample size, under
assumed fixed level of the $\pi$-estimator's variance. As in the case of the
classical optimal allocation, the problem of minimum sample size allocation can
be complemented by imposing upper bounds constraints on sample sizes in
strata.

*Stratallo* provides two user functions, `dopt` and `nopt` that solve optimal
sample allocation problems briefly characterized above. In this context, it is
assumed that the sampling designs in strata are chosen so that the variance of
the $\pi$-estimator of the population total is of the following generic form:
$$
  D^2_{st}(x_w,\, w \in \W) = \sum_{w \in \W}\, \frac{a_w^2}{x_w} - b,
$$
where $\W = \{1, \ldots, H\}$ denotes set of strata labels with total number of
strata equals to $H$, $(x_w)_{w \in \W}$ are the strata sample sizes, and
parameters $b$, and $a_w > 0,\, w \in \W$, do not depend on the
$(x_w)_{w \in \W}$. Among the stratified sampling designs that have the
$\pi$-estimator's variance of the above form is stratified simple random
sampling without replacement design. Under this design
$a_w = N_w S_w,\, w \in \W$ and $b = \sum_{w \in \W}\, N_w S_w^2$, where
$S_w,\, w \in \W$, denote stratum standard deviations of study variable and
$N_w,\, w \in \W$, are the strata sizes (see e.g. @sarndal, Result 3.7.2, p.
103).

Apart from `dopt` and `nopt`, *stratallo* provides `var_tst` and `var_tst_si`
functions that compute a value of variance $D^2_{st}$. The `var_tst_si` is a
simple wrapper of `var_tst` that is dedicated for the case of simple random
sampling without replacement design in each stratum. Furthermore, the package
comes with two predefined, artificial populations with 507 and 969 strata. These
are stored in `pop507` and `pop969` objects respectively.

## Minimization of the variance with `dopt` function

The `dopt` function solves the following two types of the allocation problem, 
formulated in the language of mathematical optimization.

**Problem 1 (one-sided upper bounds constraints)**   
Given numbers $a_w > 0,\, M_w > 0,\, w \in \W$ and $b,\, n < \sum_{w \in \W}\, M_w$,
\begin{align*}
	\underset{\x \in (0, +\infty)^{H}}{\texteq{minimize ~\,}} & \quad f(\x) = \sum_{w \in \W} \tfrac{a_w^2}{x_w} - b \\
	\texteq{subject ~ to} & \quad \sum_{w \in \W} x_w = n \\
	& \quad x_w \le M_w, \quad \forall w \in \W,
\end{align*}
where $\x = (x_w)_{w \in \W}$ is the optimization variable.

**Problem 2 (one-sided lower bounds constraints)**   
Given numbers $a_w > 0,\, m_w > 0,\, w \in \W$, and $b,\, n > \sum_{w \in \W} m_w$,
\begin{align*}
	\underset{\x \in (0, +\infty)^{H}}{\texteq{minimize ~\,}} & \quad f(\x) = \sum_{w \in \W} \tfrac{a_w^2}{x_w} - b \\
	\texteq{subject ~ to} & \quad \sum_{w \in \W} x_w = n \\
	& \quad x_w \geq m_w, \quad \forall w \in \W,
\end{align*}
where $\x = (x_w)_{w \in \W}$ is the optimization variable. 

User of `dopt` can choose whether the solution computed will be for 
**Problem 1** or **Problem 2** This is achieved with the proper use of `m` and
`M` arguments of the function. In case of **Problem 1**, user provides the
values of upper bounds with `M` argument, while leaving `m` as `NULL`. 
Similarly, for **Problem 2**, user provides the values of lower bounds with `m`
argument, while leaving `M` as `NULL`. If both `m` and `M` are specified as
`NULL` (default), the `dopt` returns the value of Tschuprov-Neyman allocation
that minimizes variance $D^2_{st}$ under constraints on total sample size
$\sum_{w \in \W} x_w = n$, and it is given by
$$
  x_w = a_w \frac{n}{\sum_{w \in \W} a_w}, \quad w \in \W
$$
There are four different algorithms available to use for **Problem 1**,
*rNa* (default), *sga*, *sgaplus*, *coma*. All these algorithms, except
*sgaplus*, are described in detail in @wesolowski2021. The *sgaplus* is defined
in @wojciak2019 as *Sequential Allocation (version 1)* algorithm.

The optimization **Problem 2** is solved by the *LrNa* that in principle is
based on the *rNa* and it is introduced in @wojciak2022.

## Minimization of the total sample size `nopt` function

The `nopt` function solves the following minimum sample size allocation problem, 
formulated in the language of mathematical optimization.

**Problem 3**  
Given numbers $a_w > 0,\, M_w > 0,\, w \in \W$, and
$b,\, D > \sum_{w \in \W} \tfrac{a_w^2}{M_w} - b > 0$,
\begin{align*}
	\underset{\x \in (0, +\infty)^{H}}{\texteq{minimize ~\,}} & \quad n(\x) = \sum_{w \in \W} x_w \\
	\texteq{subject ~ to} & \quad \sum_{w \in \W} \tfrac{a_w^2}{x_w} - b = D \\
	& \quad x_w \le M_w, \quad \forall w \in \W,
\end{align*}
where $\x = (x_w)_{w \in \W}$ is the optimization variable.

The algorithm that solves **Problem 3** is based on the *LrNa* and it is
described in @wojciak2022.

## Installation

You can install the released version of *stratallo* package from
[CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("stratallo")
```

## Examples

These are basic examples that show how to use `dopt` and `nopt` functions
to solve optimal sample allocation problems for an example population with 4
strata.

```{r load_package}
library(stratallo)
```

### Function `dopt`

```{r example_dopt_pop}
# Define example population
N <- c(3000, 4000, 5000, 2000) # Strata sizes.
S <- c(48, 79, 76, 17) # Standard deviations of a study variable in strata.
a <- N * S
```

```{r example_dopt_M}
M <- c(100, 90, 70, 80) # Upper bounds constraints imposed on the sample sizes in strata.
all(M <= N)
n <- 190 # Total sample size.
n < sum(M)

# Solution to Problem 1.
opt <- dopt(n = n, a = a, M = M)
opt
sum(opt) # Equals to n.
all(opt <= M) # Does not violate upper bounds constraints.
# Variance of the pi-estimator that corresponds to a given optimal allocation.
var_tst_si(opt, N, S)
```

```{r example_dopt_m}

m <- c(50, 120, 1, 1) # Lower bounds constraints imposed on the sample sizes in strata.
n > sum(m)

# Solution to Problem 2.
opt <- dopt(n = n, a = a, m = m)
opt
sum(opt) # Equals to n.
all(opt >= m) # Does not violate lower bounds constraints.
# Variance of the pi-estimator that corresponds to a given optimal allocation.
var_tst_si(opt, N, S)
```

```{r example_dopt_Neyman}

# Tschuprov-Neyman allocation (no inequality constraints).
opt <- dopt(n = n, a = a)
opt
sum(opt) # Equals to n.
# Variance of the pi-estimator that corresponds to a given optimal allocation.
var_tst_si(opt, N, S)
```

### Function `nopt`

```{r example_nopt}

a <- c(3000, 4000, 5000, 2000)
b <- 70000
M <- c(100, 90, 70, 80)
D <- 1e6 # Variance constraint.

opt <- nopt(D, a, b, M)
sum(opt)
```

## References
